Comparison of approaches to using machine learning models to increase the efficiency of the global search algorithm for solving multicriterial problems

### 2024-06-29 2024/OPTIMA/OPTIMA_2024_MCO.tex
OPTIMA last part

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{cite}
\usepackage{hyperref}

@@

\cite{Miettinen1999,Ehrgott2005,Pardalos2017,ML_MCO_2023,Evtushenko2014,Deb2002,Durillo2010,Mostaghim2007,NDG09,RC05,ZLT01,Gergel2019_2,Gergel2018,GergelKozinov2020,Marler2004,Strongin2000,Sergeyev2013,SVM_2000,PROB_2004,iOpt_url,Grishagin2015_2}

@@

\begin{equation}
    \label{eq:19}
		\begin{matrix}
		  f(y)= -(AB + CD)^{1/2} \\
			AB =(\sum_{i=1}^7{\sum_{j=1}^7{[A_{ij} a_{ij} (y_1,y_2) + B_{ij} b_{ij} (y_1,y_2)])}})^2 \\
			CD =(\sum_{i=1}^7{\sum_{j=1}^7{[C_{ij} a_{ij} (y_1,y_2) - D_{ij} b_{ij} (y_1,y_2)])}})^2 \\
			a_{ij} (y_1,y_2) = \sin(\pi i y_1) \sin(\pi j y_2), \\
			b_{ij} (y_1,y_2) = \cos(\pi i y_1) \cos(\pi j y_2),
		\end{matrix}
\end{equation}

\begin{figure}
\includegraphics[width=\textwidth]{fig3.png}
\label{fig:3}
\caption{A figure caption.} 
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{fig4.png}
\label{fig:4}
\caption{A figure caption.} 
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{fig5.png}
\label{fig:5}
\caption{A figure caption.} 
\end{figure}

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{cite}
\usepackage{hyperref}

@@

@@

Computations were conducted on the supercomputer ``Lobachevsky'' of the Nizhni Novgorod State University (operating system CentOS 7, control system SLURM). To obtain the executable program code the compilers Intel C++ 17.0.0 and python 3.9 were used. Computations were performed with the support of the package scikit-learn 0.24.2 and the optimization software iOpt \cite{iOpt_url}.

In the first experiment 100 bi-criterial MCO problems have been solved. As the partial criteria, the functions \cite{ML_MCO_2023} 
\begin{equation}
    \label{eq:19}
		\begin{matrix}
		  f(y)= -(AB + CD)^{1/2} \\
			AB =(\sum_{i=1}^7{\sum_{j=1}^7{[A_{ij} a_{ij} (y_1,y_2) + B_{ij} b_{ij} (y_1,y_2)])}})^2 \\
			CD =(\sum_{i=1}^7{\sum_{j=1}^7{[C_{ij} a_{ij} (y_1,y_2) - D_{ij} b_{ij} (y_1,y_2)])}})^2 \\
			a_{ij} (y_1,y_2) = \sin(\pi i y_1) \sin(\pi j y_2), \\
			b_{ij} (y_1,y_2) = \cos(\pi i y_1) \cos(\pi j y_2),
		\end{matrix}
\end{equation}
were taken  where $0 \leq y_1, y_2 \leq 1$, parameters $-1 \leq A_{ij},B_{ij},C_{ij},D_{ij} \leq 1$ are independent uniformly distributed random values.

Computational experiments for techniques of Subsection \ref{ssec:42} which utilize probabilities of belonging to the Pareto set can be divided into 3 groups.
\begin{enumerate}
	\item 	Experiments with the algorithm ML\_MGSA\_Prob (fixed weights).
	\item 	Solving MCO problems by the method ML\_MGSA\_Prob (Adjustive weights));
	\item 	Experiments with changing weights in the course of ML\_MGSA\_Prob running and transformed probabilities (\ref{eq:18})  (ML\_MGSA\_Prob (Norm Log)).
\end{enumerate}

In each series of experiments for these optimization algorithms, the coefficients $\alpha$ from (\ref{eq:16}) changed in the range $[0.1, 0.9]$. Comparison with the algorithm MGSA that does not use machine learning methods (in the characteristic ($\alpha$) $\alpha = 0$) and with the algorithm ML\_MGSA\_Dist taking into account the distance to the separating hyperplane was carried out. The comparison was made on the base of  the calculated HV index characterizing quality of the obtained Pareto set approximation \cite{ML_MCO_2023,Evtushenko2014}. Further subsections include the results of only those calculations in which the $\alpha$ parameter provided the best quality of the solution found. Tables contain results of solving all the problems (\ref{eq:07}) with fixed accuracy $\varepsilon=0.01$ in the termination condition (\ref{eq:14}). Figures present results for a series of decreasing $\varepsilon$ which generates increasing numbers of trials.

\textbf{Experiments with fixed weights of two trial classes.} The first series of experiments is aimed at comparison the algorithm ML\_MGSA\_Prob (Fixed weights) with results of running for methods MGSA and ML\_MGSA\_Dist.

The following parameters are indicated in Table \ref{tab:1}: the kernel of the SVM model (column 2), parameter $\alpha$ (column 3), average values of the HV-index (column 4), the average number of trials spent by the algorithm (column 5), and the coefficient of decreasing the number of trials compared to ML\_MGSA (column 6). The factors $r$ and $\varepsilon$ from (\ref{eq:10}) and (\ref{eq:14}) were the same for all the algorithms.

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[ht]
\centering
\caption{Results of computational experiments}
\label{tab:1}
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\hline
\textbf{}                                                                                  & \textbf{Model Kernel} & \textbf{$\alpha$} & \textbf{Avg. HV-index} & \textbf{Avg. iter.} & \textbf{Speedup iter.} \\ \hline
ML\_MGSA                                                                                   & -                     & -                              & 114.91                 & 2269.82             & 1.00                   \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of fixed weights}}                                                                                                                                                      \\ \hline
ML\_MGSA\_Dist                                                                             & Linear                & 0.01                           & 114.67                 & 691.14              & \textbf{3.28}          \\
\multirow{3}{*}{\begin{tabular}[c]{@{}c@{}}ML\_MGSA\_Prob \\ (Fixed weights)\end{tabular}} & Linear                & 0.04                           & 114.72                 & 963.05              & 2.36                   \\
                                                                                           & Poly                  & 0.07                           & 114.73                 & 871.82              & 2.60                   \\
                                                                                           & RBF                   & 0.03                           & 114.80                 & 883.04              & 2.57                   \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights}}                                                                                                                                                   \\ \hline
ML\_MGSA\_Dist                                                                             & Linear                & 0.01                           & 114.67                 & 691.14              & \textbf{3.28}          \\
\multirow{3}{*}{\begin{tabular}[c]{@{}c@{}}ML\_MGSA\_Prob \\ (Adj. weights)\end{tabular}}  & Linear                & 0.04                           & 114.76                 & 1018.03             & 2.23                   \\
                                                                                           & Poly                  & 0.08                           & 114.73                 & 844.66              & 2.69                   \\
                                                                                           & RBF                   & 0.03                           & 114.76                 & 796.74              & 2.85                   \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights and logarithmic probabilities}}                                                                                                                              \\ \hline
ML\_MGSA\_Dist                                                                             & Linear                & 0.01                           & 114.67                 & 691.14              & \textbf{3.28}          \\
\multirow{3}{*}{\begin{tabular}[c]{@{}c@{}}ML\_MGSA\_Prob \\ (Log Norm)\end{tabular}}      & Linear                & 0.05                           & 114.72                 & 711.44              & 3.19                   \\
                                                                                           & Poly                  & 0.04                           & 114.79                 & 854.82              & 2.66                   \\
                                                                                           & RBF                   & 0.04                           & 114.74                 & 704.99              & 3.22                   \\ \hline
\multicolumn{6}{c}{\textbf{Experiments with 4-criterial MCO problems}}                                                                                                                                                      \\ \hline
ML\_MGSA\_Dist                                                                             & Linear                & 0.01                           & 7751.43                & 1925.6              & 1.00                   \\
\multirow{3}{*}{ML\_MGSA\_Prob (Log Norm)}                                                 & Linear                & 0.08                           & 7729.71                & 1760.5              & 1.09                   \\
                                                                                           & Poly                  & 0.08                           & 7796.2                 & 4121.7              & 0.47                   \\
                                                                                           & RBF                   & 0.08                           & 7786.48                & 1364.6              & \textbf{1.41}          \\ \hline
\end{tabular}%
}
\end{table}

Fig. \ref{fig:3} shows the change of the average HV-index as a function of the number of trials.
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{fig3.png}
\caption{Comparison of HV-indices for the case of fixed weights.} 
\label{fig:3}
\end{figure}

Compared to MGSA, the other algorithms reduce significantly the number of iterations in conditions when HV-indices are similar. However, the iteration number when using the models with probabilities (see last 3 rows of Table \ref{tab:1} at section ``Experiments for the case of fixed weights'') is greater than for the algorithm based on the separating hyperplane.

\textbf{Experiments with tuned weights in the course of optimization.} In this group of experiments the algorithm ML\_MGSA\_Prob (adj. weights) changed weights of trial classes before starting a new subproblem (\ref{eq:07}). Table \ref{tab:1} and Fig. \ref{fig:4} for these experiments are organized similarly to the first series of experiments.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{fig4.png}
\caption{HV-indices comparison  in the case of adjusted weights} 
\label{fig:4}
\end{figure}

The presented data on methods using probabilities, as in the previous series of experiments, show significantly better results in the number of iterations compared to an algorithm without machine learning techniques. Nevertheless, from the results obtained, one cannot unequivocally say about a significant improvement or deterioration of these results compared to experiments in which the weights were set fixed.

\textbf{Experiments with adjusted weights and logarithmic probabilities.} In the third group of experiments, in addition to changing the weights during the algorithm’s running, the transformation (\ref{eq:18}) was used in techniques with probability calculation (methods ML\_MGSA\_Prob (Log Norm)). Table \ref{tab:1} and Fig. \ref{fig:5} are similar to those in the previous experiments.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{fig5.png}
\caption{Comparison of  HV-indices for methods with probabilities (\ref{eq:18})} 
\label{fig:5}
\end{figure}

From the results of the experiment, the following conclusions can be drawn. Among the methods applying probabilistic techniques, the method with the transformation (\ref{eq:18}) demonstrates the best results. In the case of the SVM model with the linear kernel and with the RBF kernel, the results under increasing accuracy are comparable to the results obtained for ML\_MGSA\_Dist. At the same time, the techiques with probabilities enable changing easier the machine learning models.

\textbf{Experiments with 4-criterial MCO problems.} In the last series of experiments, 10 MCO problems with 4 criteria were solved that form a Pareto set of complex structure. As in previous subsection, the functions of the family (\ref{eq:19}) were used as partial criteria. In the course of previous experiments, it was found that the closest result to ML\_MGSA\_Dist is given by the algorithm ML\_MGSA\_Prob (Log Norm) based on the technique (\ref{eq:18}). Taking into account this fact, we limited ourselves to comparing the results of these two algorithms. The results are presented in Table \ref{tab:1} and Fig. \ref{fig:6}

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{Fig6.png}
\caption{Efficiency comparison in the 4-critarial case} 
\label{fig:6}
\end{figure}

The results of the experiments allow us to conclude that the method ML\_MGSA\_Prob (Log Norm) in this series of experiments can significantly reduce the number of trials with a similar value of HV-index compared to ML\_MGSA\_Dist. The best results were achieved using models with Linear and RBF kernels, while the latter model demonstrates better quality than the version of the algorithm with the technique \cite{ML_MCO_2023} based on distances to the hyperplane.

\section{Conclusion}

The paper proposes a new modification of the global search algorithm focused on solving problems of multicriteria optimization. The algorithm in question is based on criteria scalarization and solving a series of multiextremal scalar problems.

Acceleration of the algorithm (in terms of reducing the number of search trials required to evaluate the Pareto set) is achieved through the application of machine learning methods. Namely, when using the SVM method, trial points in the criteria space are classified (each point is/is not an effective solution), on the basis of which a numerical estimate of the probability of belonging to the Pareto set is built. Polynomial and radial basis function kernels in SVM allows one to adequately approximate Pareto front of a complex structure.

Solving several series of MCO problems with complex multiextremal criteria has demonstrated that the proposed approach to evaluating the set of effective solutions is promising. Further research aims to develop the proposed approach by consideration of other machine learning models (in particular, artificial neural networks) to better classify trials and approximate the Pareto set.
